### 2024-03-19 latex/HP4_comm.tex
adapted to new input files and cleanup
new files and Nagari-Support for Jyotsna

\input{Preamble-Print}
\input{CommStyle}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

\chapter*{Translation \& philological commentary}
%%%%%%%%%%
\subsection*{3.1}
\begin{translation}[hp03_001]
test.
\end{translation}

>>>>

\input{PreambleComm}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

%\chapter*{Translation \& philological commentary}

%%%%%%%%%%
\subsection*{4.0*1}
\begin{translation}[hp04_000_1] 
Homage to the guru, Śiva, who consists of \emph{nāda}, \emph{bindu} and \emph{kalā}. [The yogi] who is constantly devoted to him attains the untainted state (\emph{nirañjanapada}).
\end{translation}

%\begin{sources}[hp04_000_1]
%\end{sources}

\begin{testimonia}[hp04_000_1]
\emph{Haṃsavilāsa} 14 (p. 47)
\begin{versinnote}
\tl{namaḥ śivāya gurave nādabindukalātmane |\\+}
\tl{nirañjanapadaṃ yāti yatra yogī parāyaṇaḥ iti dhyānam || 14 ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_000_1]
%The compound \emph{yatnaparāyaṇa} is rather unusual and generally not attested elsewhere (the only exception known  to us is the \emph{Amarasiddhi}). The reading  \emph{yogaparāyaṇa} in some secondary manuscripts makes much better sense (i.e., 'one devoted to yoga') but is likely a substitution for a more difficult reading.
%adopt yatraparāyaṇaḥ [MD: done], and assume tasmai in 4.1ab

%Look into referent of nirañjana in these texts. Judit: best to take it generally (like nirāmaya, etc.).
Verses 4.0*1–4.0*16 are omitted by the \textalpha\ group and are likely not to be original. The first additional verse resembles a \emph{maṅgala} verse that one might expect to see at the beginning of a text. The second is a verse from the \emph{Gorakṣaśataka} that introduces the topic of \emph{samadhi}. Some manuscripts of the \textbeta, \textepsilon\ and \texteta\ groups have the two verses on the synonyms of \emph{rājayoga} here (on their position in the \textalpha\ group and other manuscripts, see the note to 4.29). The rest of the additional verses (4.0*5–4.0*16) are a motley collection on \emph{samādhi}, \emph{rājayoga}, the importance of the guru, dissolving the breath, \emph{suṣumṇā}, etc. In contrast to this, the \textalpha\ group begins with a cohesive discusson on absorption (4.1–3) that transitions to the gaze (4.4) and a brief discussion of \emph{śāmbhavī} and \emph{khecarī mudrā}s (4.5–4.8). The main topic of the chapter, which is meditating on the internal sound (\emph{nādānusandhāna}), begins at 4.9 in the \textalpha\ group (whereas in other groups it begins after fifty or so verses). The emphasis on \emph{nādānusandhāna} in the fourth chapter of the \textalpha\ group is consistent with the statement in verse 1.56 that \emph{nādānusandhāna} is the fourth component of Haṭhayoga.

Its likely that the term \emph{nirañjanapada} was understood here as \emph{samādhi} because \emph{nirañjana} is included in a list of synonyms of \emph{samādhi} later in this chapter (4.29).

The triad \emph{nāda}, \emph{bindu} and \emph{kalā} occurs in earlier works, in particular Śaiva Tantras, where it appears in contexts of enunciating mantras (\emph{mantroccāra}, e.g., \emph{Kubjikāmatatantra} 7.65, \emph{Jñānārṇavatantra} 2.4, \emph{Īśānaśivagurudevapaddhati} 18-110, etc.) and sometimes qualifies dieties (e.g., \emph{Parākhyatantra} 5.156ab) and gurus (e.g., \emph{Gurugītā} 64). The context can change the meaning of these terms, so we have chosen not to translate them. For a discussion of their various meanings, see \emph{Tāntrikābhidhānakośa} 2004 vol. 2: 68–73, 2013 vol. 3, 277–279.

%nāda-bindu-kalā, referring to inner sound, generative fluid, and digits of the moon (or unmanīkalā) in the context of HY? Judit: kalā is like śakti (initial source)?
%Refer to TĀK

\end{philcomm}

\begin{metre}[hp04_000_1]
Anuṣṭubh (a: na-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{4.0*2}
\begin{translation}[hp04_000_2]
So now I will teach the best way to \emph{samādhi}.It destroys death, has an easy method and brings about the bliss of Brahman.
\end{translation}

\begin{sources}[hp04_000_2]
\emph{Gorakṣaśataka} 64
\begin{versinnote}
\tl{athedānīṃ pravakṣyāmi samādhikramam uttamam | \\+}
\tl{mṛtyughnaṃ tu sukhopāyair brahmānandakaraṃ sadā ||\\+}
\tl{\var{64c tu sukhopāyair ] T; sukhadopāyaṃ GU}\\!}
\end{versinnote}
\end{sources}

%\begin{testimonia}[hp04_000_2]
%\end{testimonia}

%\begin{philcomm}[hp04_000_2]
%It appears that the \emph{Gorakṣaśataka}'s verse was changed by whoever added it to reflect the discursive nature of the discussion on \emph{samādhi} that follows. By changing \emph{samādhikramam uttamam} to \emph{samādhikramalakṣaṇam} (and perhaps \emph{sadā} to \emph{param}) the verse foreshadows a broader discussion of the topic samādhi. But it doesn't really work, so adopt uttamam. [JB: We have uttamam now, so this note is not needed]
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*5}
\begin{translation}[hp04_000_5]
The unity of the self and mind arises in the same way that salt becomes identical with water through contact [with it]. That is called \emph{samādhi}.
\end{translation}

\begin{sources}[hp04_000_5]
\emph{Vivekamārtaṇḍa} 161
\begin{versinnote}
\tl{ambusaindhavayoḥ sāmyaṃ yathā bhavati yogataḥ |\\+}
\tl{tathātmamanasor aikyaṃ samādhiḥ so’bhidhīyate ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_000_5]
\emph{Haṭharatnāvalī} 4.1
\begin{versinnote}
\tl{salile saindhavaṃ yadvat sāmyaṃ bhavati yogavit |\\+}
\tl{tathātmamanasor aikyaṃ samādhiḥ so 'bhidhīyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 11.29 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{ambusaindhavayor aikyaṃ yathā bhavati yogataḥ |\\+}
\tl{tathātmamanasor aikyaṃ samādhir abhidhīyate ||\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 51.72
\begin{versinnote}
\tl{tad uktaṃ yogacandrikāyām –\\+}
\tl{ambusaindhavayor aikyaṃ yathā bhavati yogataḥ |\\+}
\tl{tathātmanasor aikyaṃ samādhiḥ sa vidhīyate ||\\+}
\tl{aikyaṃ abhinnatvam | yogataḥ yogābhyāsāt ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_000_5]
%The change \emph{salile saindhavaṃ} appears to have occured because of the verb \emph{bhajati}, which is different to the \emph{Vivekamartaṇḍa}'s reading of \emph{ambusaindhavayor aikyaṃ [...] bhavati [...]}. However, since groups 2 and 3 support \emph{salile}, it is possible that Svātmārāma accepted it.
%SS sees this verse as a nirukti of samādhi using sāmyam
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*6}
\begin{translation}[hp04_000_6]
The identity of the individual and universal selves is called \emph{samādhi}, in which all thoughts disappear.
\end{translation}

\begin{sources}[hp04_000_6]
\emph{Vivekamārtaṇḍa} 163
\begin{versinnote}
\tl{yat samatvaṃ dvayor atra jīvātmaparamātmanoḥ |\\+}
\tl{samastanaṣṭasaṃkalpaḥ samādhiḥ so’bhidhīyate ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_000_6]
\emph{Haṭharatnāvalī} 4.2
\begin{versinnote}
\tl{tat samatvaṃ bhaved atra jīvātmaparamātmanoḥ |\\+}
\tl{samastanaṣṭasaṃkalpaḥ samādhiḥ so 'bhidhīyate ||\\!}
\end{versinnote}

\emph{Yuktabhavadeva} 11.28 (attr.~to Gorakṣanātha)
\begin{versinnote}
\tl{yat sarvadvandvayor aikyaṃ jīvātmaparamātmanoḥ |\\+}
\tl{samastanaṣṭasaṃkalpaḥ samādhiḥ so 'bhidhīyate ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp04_000_6]
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*7}
\begin{translation}[hp04_000_7]
Who indeed truly knows the majesty of Rājayoga? From knowledge, liberation †arises† [and] power (\emph{siddhi}) is obtained by means of the guru's teaching.
\end{translation}

\begin{sources}[hp04_000_7]
\emph{Amanaska} 2.5
\begin{versinnote}
\tl{rājayogasya māhātmyaṃ ko vā jānāti tattvataḥ |\\+}
\tl{jñānāt siddhir muktir iti guror jñānaṃ ca labhyate ||\\+}
\tl{\var{5c jñānāt siddhir muktir iti ] Bl Jb K Ma Va:
jñānāt siddhimuktir iti Pa Pc Tr Ua Ea:
jñānasiddhir muktir iti Vb Vd:
jñānān mukteḥ siddhir iti Ja:
jñānāt sidhyati muktir hi AllN (except Na Nm Ve Ea):
jñānāt sidhyate muktir hi Ve:
jñānāt sidhyanti muktiṃ hi Na:
jñānāt sidhyati muktiṃ hi Nm:
nānāsiddhir muktir iti Tha:
jñānasiddhir bhavaty eva Mb:
jñānasiddhimuktisiddhi Pb:
jñānāt siddhimuktor iti Je}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_000_7]
\emph{Yogacintāmaṇi} f.~37v (citing the \emph{Rājayoga} [aka \emph{Amanaska}])
\begin{versinnote}
\tl{rājayogasya māhātmyaṃ ko hi jānāti tattvataḥ |\\+}
\tl{tajjñānī vasate yatra sadeśaḥ puṇyabhājanam ||\\!}
\end{versinnote}

%\emph{Yogamārgaprakāśikā} 4.11
%\begin{versinnote}
%\tl{rājayogasya māhātmyaṃ ko vā śaknoti varṇitum | \\+}
%\tl{yogasyāsya ca kartāro vijñeyās te maheśvarāḥ ||\\!}
%\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_000_7]
The third verse quarter of this verse has been subject to much revision in the \emph{Haṭhapradīpikā} and the source text, the \emph{Amanaska}. Unlike the manuscripts of the \emph{Amanaska}, those of the \emph{Haṭhapradīpikā} transmit \emph{sthitiḥ} or \emph{sthite} after \emph{muktiḥ} or \emph{mukti}. We are unable to make sense of \emph{sthiti} here, and so have cruxed this verse quarter and emended to \emph{muktiḥ sthitā} (`liberation has happened').
%?? JB. Rather than emend to muktiḥ sthitā, I think we should consider muktisthitiḥ (etaTwo and deltaOne), which is similar to jīvanmuktasthitiṃ in the Mokṣopāya, LYV, MSS, etc. We could translate it as Sternbach below ('state of liberation'). JM: I also don't think we should emend. muktisthitiḥ is possible (but MU parallel not great because of mukti vs mukta). We could also consider muktiḥ sthitiḥ “liberation [and] stability”
% Mokṣopāya
% jīvanmuktasthitiṃ brahman kṛtvā rāghavam āditaḥ|\\
% kramāt kathaya me nityaṃ bhaviṣyāmi sukhī yathā||\,1||
%  "jīvanmuktasthitiṃ" jīvanmuktamaryādāṃ
% Yogavāsiṣṭha
% vicārajā nijā jīvanmuktasthitir ihoditā |
% muninābhyāsahīnasya vyādhasya tvanavasthitiḥ || VII. 
% MSS_9744-1 
% kāryākārye kim api satataṃ naiva kartṛtvam asti  
% jīvanmuktasthitir avagato dagdhavastrāvabhāsaḥ /<BR>
% There is not always an agent in every action, good or bad ; for, the state of a jivanmukta [one who is released from bondage while living in the world] is considered to be similar to that of the splendour of a burnt cloth [which can neither be clean nor dirty]. (Sternbach's translation)
\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*8}
\begin{translation}[hp04_000_8]
Letting go of sense objects, seeing the truth, [and] realising the innate state are difficult without the compassion of a good guru.
\end{translation}

%\begin{sources}[hp04_000_8]
%\end{sources}

%\begin{testimonia}[hp04_000_8]
%\end{testimonia}

%\begin{philcomm}[hp04_000_8]
%No source?
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*9}
\begin{translation}[hp04_000_9]
When the primal \emph{śakti} (i.e.~\emph{kuṇḍalinī}) has been woken up by means of the various postures, retentions and wonderful techniques [i.e.~\emph{mudrā}s], the breath dissolves into the void.
\end{translation}

%\begin{sources}[hp04_000_9]
%\end{sources}

\begin{testimonia}[hp04_000_9]
\emph{Yogacintāmaṇi}
\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{vividhair āsanaiḥ kumbhair vicitrakaraṇair api |\\+}
\tl{prabuddhāyām ādiśaktau prāṇaḥ śūnye vilīyate ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_000_9]
In \emph{Jyotsnā} 4.10, Brahmānanda understands `the void' (\emph{śūnya}) as the central channel. In \emph{Haṭhapradīpikā} 3.4, \emph{śūnyapādavī} is a synonym of \emph{suṣumnā}.
\end{philcomm}
%?? MD: vilīyate or pralīyate?

\begin{metre}[hp04_000_9]
Anuṣṭubh (c: ra-vipulā)
\end{metre}

%%%%%%%%%%
\subsection*{4.0*10}
\begin{translation}[hp04_000_10]
For the yogi whose \emph{kuṇḍalinī} has awakened and who has given up all activity, the innate state automatically arises.
\end{translation}

%\begin{sources}[hp04_000_10]
%\end{sources}

\begin{testimonia}[hp04_000_10]
\emph{Yogacintāmaṇi} f.~9r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{utpannaśaktibodhasya tyaktaniḥśeṣakarmaṇaḥ |\\+}
\tl{yoginaḥ sahajāvasthā svayam eva prajāyate ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_000_10]
%\end{philcomm}

%?? MD: prajāyate or prakāśate?

%%%%%%%%%%
\subsection*{4.0*11}
\begin{translation}[hp04_000_11]
When the breath is flowing in the central channel and the mind enters the void, the knower of yoga uproots all [past] actions.
\end{translation}
%?? JB: all [fruits of past] actions

\begin{sources}[hp04_000_11]
\end{sources}

\begin{testimonia}[hp04_000_11]
\emph{Yogacintāmaṇi} f.~9r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{suṣumṇāvāhini prāṇe śūnye viśati mārute |\\+}
\tl{tathā samastakarmāṇi nirmūlayati yogavit ||\\!}
\end{versinnote}

\emph{Upāsanāsārasaṅgraha} p.~66 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{suṣumṇāvāhini prāṇe śūnye viśati mānase |\\+}
\tl{tadā samastakarmāṇi nirmūlayati marmavit ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_000_11]
%\emph{tathā} may connect the two locative absolutes.
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*12}
\begin{translation}[hp04_000_12]
Homage to you, the immortal one. You have slain even death, into whose mouth this world, with everything that is moving and unmoving, has fallen.
\end{translation}

\begin{sources}[hp04_000_12]
\end{sources}

\begin{testimonia}[hp04_000_12]
\end{testimonia}

%\begin{philcomm}[hp04_000_12]
%It appears that this verse, which is in manuscripts of the \textbeta\ and \textepsilon\ groups, has become corrupt in manuscripts of \textgamma\, perhaps because of eye-skip (i.e., \emph{amano nirmanāḥ śūnyaṃ jagad etac carācaram}).
%?? JB should we include this note on gamma's omission? Is it necessary? JM: no, I've percented it

%JT: alt 3 is original, makes best sense, others lost pādas b and c and tried to correct. Adopt. [MD: done]
%Eye skip might explain how amarāya namas became amano nirmanaḥ śūnyaṃ jagad ... and the tadāmaraugha was dropped because it was difficult to understand or was corrupt.
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*13}
\begin{translation}[hp04_000_13]
When the mind has obtained equanimity and the breath is moving into the central channel, then the \emph{vajrolī} of the lineage of immortals arises even for the person whose hope is for that.
\end{translation}
% ?? JB: we didnt translate this verse in the workshop. I've adapted my translation from a similar verse in the Amaraugha. JM: equanimity is tautologous with mind, “When equanimity has been obtained"? Also I've changed tr. to reflect tadā.

\begin{sources}[hp04_000_13]
\emph{Amaraugha} 7
\begin{versinnote}
\tl{citte tu sattvam āpanne vāyau vrajati madhyame |\\+}
\tl{tadāmaraughavajrolis tadāśājīvitasya ca ||\\+}
\tl{\var{citte tu sattvam ] $\beta$ : citte samatvam Ba Ad Ed : cittau samatvam Gb Ta.
tadāmaraughavajrolis ] \textit{conj.} :
tadāmaraughavajroli Ga :
tadāmaraudhavajrolī Ae :
eṣāmarauli vajroli Gb :
eṣāmarāli vajroli Ba :
eṣāmarauli vajrauli Ta :
yeṣāmarāli vajrauli Ad :
eṣāmarolī vajrolī Ed.
tadāśājīvitasya ca ] $\beta$ :
tadā meni mameti ca Ba :
tadā mati mateti ca Ed :
mameti ca mameti ca Ad :
tadā meti mameti mateti ca Gb (\textit{unmetr.}) :
mateti mame++ Ta.
}\\!}
\end{versinnote}

\end{sources}

%\begin{testimonia}[hp04_000_13]
%\end{testimonia}

\begin{philcomm}[hp04_000_13]
The second line of this verse is problematic in both the source text (the \emph{Amaraugha}) and the manuscripts of the \emph{Haṭhapradīpikā}. It seems that the \textepsilon\ group has preserved the most plausible version of this line, which is also quite close to the source text. The line was rewritten in \etaTwo\ and other manuscripts, as well as the \emph{Jyotsnā} (4.14cd), to include the two variations of \emph{vajrolī}, namely \emph{amarolī} and \emph{sahajolī}.%??JM: comment on pāda d? It's odd, worth cruxing even.
%C2 (Grp 7) tadāmarolī ... sahajolī prajāyate (=Jyotsnā)
\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*14}
\begin{translation}[hp04_000_14]
How can knowledge exist, when the mind is alive, O goddess? So long as the breath lives, the mind does not die. Breath and mind: he who leads this pair to dissolution attains liberation; in no way [does any] other man.
\end{translation}

\begin{sources}[hp04_000_14]
\emph{Candrāvalokana} 6
\begin{versinnote}
\tl{jñānaṃ kuto manasi jīvati devi yāvat \\+}
\tl{prāṇo pi jīvati mano mriyate na yāvat |\\+}
\tl{prāṇo mano dvayam idaṃ vilayaṃ prayāti\\+}
\tl{mokṣaṃ sa gacchati naro na kathaṃcid anyaḥ ||\\+}
\tl{\var{devi ] 7970 4340 T00788, kūdṛṣṭi D4345 (unmetr.)}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_000_14]
\emph{Yogacintāmaṇi} f.~17r
\begin{versinnote}
\tl{atra śivavākyam |\\+}
\tl{jñānaṃ kuto manasi jīvati devi tāvat\\+}
\tl{prāṇo 'pi jīvati mano mriyate na yāvat |\\+}
\tl{prāṇo mano dvayam idaṃ vilayaṃ nayed yo \\+}
\tl{mokṣaṃ sa gacchati naro na kathañcid anyaḥ ||\\!}
\end{versinnote}

\emph{Haṭhasaṅketacandrikā} f.~2r
\begin{versinnote}
\tl{jñānaṃ kuto manasi jīvati durvikalpe \\+}
\tl{prāṇe 'pi jīvati mano mṛyate na yāvat |\\+}
\tl{prāṇo mano dvayam idaṃ na vilīyate 'tra \\+}
\tl{mokṣa na gacchati naro 'tra kathaṃ cid eva ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_000_14]
%\end{philcomm}
%?? MD: adopt prayāti (beta) in Pāda C?

\begin{metre}[hp04_000_14]
Vasantatilakā 
\end{metre}

%%%%%%%%%%
\subsection*{4.0*15}
\begin{translation}[hp04_000_15]
Both mercury and the mind are unsteady by nature. [When either] mercury is fixed [or] the mind is fixed, nothing in the world is impossible.
\end{translation}

%\begin{sources}[hp04_000_15]
%\end{sources}

\begin{testimonia}[hp04_000_15]
\emph{Yogacintāmaṇi} f.~103v
\begin{versinnote}
\tl{śāstrāntare śivavākyam—\\+}
\tl{rasasya manasaś caiva cañcalatvaṃ samāsataḥ |\\+}
\tl{raso baddho mano baddhaṃ kiṃ na sidhyati bhūtale ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_000_15]
%LO: rasa implies bindu (and the next verse is about bindu when in the AS). But the next verse is more likely from the Rasārṇava (look at the context there).
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.0*16}
\begin{translation}[hp04_000_16]
Stabilised, mercury and breath cure disease; stilled, they automatically bring back life; and bound they bestow the ability to fly, O Bhairavī.
\end{translation}

\begin{sources}[hp04_000_16]
\emph{Rasārṇava} 1.19
\begin{versinnote}
\tl{mūrchito harati vyādhiṃ mṛto jīvayati svayam |\\+}
\tl{baddhaḥ khecaratāṃ kuryāt raso vāyuś ca bhairavi ||\\!}
\end{versinnote}

Cf.~\emph{Amṛtasiddhi} 7.7
\begin{versinnote}
\tl{mūrcchito harate vyādhiṃ baddhaḥ khecaratāṃ nayet |\\+}
\tl{sarvasiddhikaro līno niścalo muktidāyakaḥ ||\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_000_16]
\emph{Yogacintāmaṇi} f.~103v (attr.~to \emph{śāstrāntare śivavākyam})
\begin{versinnote}
\tl{mūrchito harate vyādhiṃ mṛto jīvayati svayam |\\+}
\tl{baddhaḥ khecaratāṃ dhatte mano vāyuś ca bhairavi ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_000_16]
The \emph{Rasārṇava} is the likely source of this verse given the vocative, \emph{bhairavi}. Similar verses are found in other yoga texts (see Mallinson and Szántó 2021: 57).

Disregarding the technical meaning of the terms, the verse plays with apparent contradictions (\emph{virodhābhāsa}) in saying literally that, when mercury and the breath are unconscious, they cure disease; when dead, they restore life and, when bound, they enable one to fly up.
\end{philcomm}

%%%%%%%%%%
\subsection*{4.1}
\begin{translation}[hp04_001]
%Now, \emph{samādhi}–
Mind is the master of the senses; breath is the master of the mind [and] dissolution [of the mind] is the master of the breath. [The yogi] should take refuge in that master, dissolution [of the mind].
\end{translation}
%?? JB: Both \alphaOne and \alphaTwo have atha samādhiḥ as a heading before this verse. Shouldnt we include this in our edition or at least note it in the apparatus? We've added atha headings elsewhere from alpha. There's also the issue of where to place atha samādhi in relation to the greyscale verses.

%\begin{sources}[hp04_001]
%\end{sources}

\begin{testimonia}[hp04_001]
\emph{Haṭharatnāvalī} 4.4
\begin{versinnote}
\tl{indriyāṇāṃ mano nātho manonāthas tu mārutaḥ |\\+}
\tl{mārutasya layo nāthaḥ sa layo nādam āśritaḥ ||\\!}
\end{versinnote}

\emph{Yogacintāmaṇi} f.~23r
\begin{versinnote}
\tl{haṭhapradīpikāyām—\\+}
\tl{indriyāṇāṃ mano nātho manonāthaś ca mārutaḥ |\\+}
\tl{mārutasya layo nāthaḥ sa layo nādam āśritaḥ ||\\!}
\end{versinnote}

\end{testimonia}

%\begin{philcomm}[hp04_001]
%
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.1*1}
\begin{translation}[hp04_001_1]
Whether or not this [dissolution] is called liberation  in another teaching, there is no difference in the bliss arising from the dissolution of mind and breath.
\end{translation}
%?? JB: 'in another doctrine' for matāntare?
%?? JB: extraordinary bliss (kaścid?)
%\begin{sources}[hp04_001_1]
%\end{sources}

\begin{testimonia}[hp04_001_1]
\emph{Yogacintāmaṇi} f.~23r (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{ayam eva tu mokṣākhyo astu vāpi matāntare |\\+}
\tl{manaḥprāṇalayo nādo na ca kaś cit vibhidyate ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_001_1]
Some of the \textbeta, \textepsilon\ and \texteta\ groups have a different version of the second half of this verse, the most plausibe of which is, 'the extraordinary bliss of dissolution of mind and breath arises in me' (\emph{manaḥprāṇalayānando mayi kaścit pravartate}). The first person pronoun suggests that this verse derives from a dialogue of some kind before it was changed to the version in the \textgamma\ and \textdelta\ manuscripts.  
\end{philcomm}

% ?? MD: comment on the variant reading "manaḥprāṇalayānando mayi kaścit pravartate"?
% JB: maybe this is the lectio difficilor and the first person pronoun was later written out.

%%%%%%%%%%
\subsection*{4.2}
\begin{translation}[hp04_002]
A motionless and unchanging dissolution [of mind], in which inhalation and exhalation have disappeared [and] perception of sense objects has ceased, is supreme for yogis.%??JM: or "The yogis’ motionless and unchanging dissolution... is the best."
\end{translation}
% on the translation of jayati as an intransitive verb, see Alexis' footnote on jayati in his forthcoming book on the Tantrāloka āhnika 1

\begin{sources}[hp04_002]
\emph{Amanaska} 2.21
\begin{versinnote}
\tl{pranaṣṭocchvāsaniśvāsaḥ pradhvastaviṣayagrahaḥ |\\+}
\tl{niśceṣṭo nirgatārambho hy ānandaṃ yāti yogavit ||\\+}
\tl{\var{pranaṣṭocchvāsa°  ] Jb Pa Va Bl Ja Ad AllSI  AllN, 
pranaṣṭasvāsa°  Mb,
pranaṣṭāsvāsa° Ua,
praṇaśocchvāsa° Ma,
pranaṣṭaḥ svāsa° Je Vb,
pranaṣṭo śvāsa°ḥ Pc,
prāṇastho śvāsa° Pb,
pranaṣṭe svāsa° K}\\!} % retained because we've read against alpha.??jM: simplify? It's only the ucch that's different. JB: done.
%\tl{\var{hy ānandaṃ yāti yogavit ]  AllNI (except  Ma Pa Pb Pc Tr Va),
%hy ānandaṃ yāti tattvavit Pa Tr Va Ea,
%svānandaṃ yāti yogavit Vd,
%hy ānandayati yoginaḥ Ma Tha,
%hy ānanday[a]ti yogī kaḥ Pc,
%hy ānanda iva yoginaḥ AllN (except  Na Nm Ea Ve),
%hy ānanda iva yogi Na Nm,
%hy ānandaś ca yoginaḥ Ve,
%hy ānandasukhayogavit Pb,
%layo jayati yoginaḥ Ad,
%layo jayati yoginām Cc}\\!} %retained to show that jayati was in the South Indian transmissin of the Amanaska.??JM: but it's in almost all HP witnesses too so perhaps variants not needed here, just mention it in comm? Or simplify? JB done
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_002]
\emph{Yogacintāmaṇi} f.~27v (attr.~to the \textit{Rājayoga})
\begin{versinnote}
\tl{praṇaṣṭocchvāsaniḥśvāsa[ḥ] vidhvastaviṣayagrahaḥ |\\+}
\tl{niśceṣṭo nirgatārambho hy ānandayati yogikaḥ ||\\!}
\end{versinnote}

\emph{Haṭhasaṅketacandrikā} f.~117 (attr.~to the \emph{Haṭhapradīpikā})
\begin{versinnote}
\tl{pranaṣṭaśvāsaniḥśvāsaḥ pradhvastaviṣayajvaraḥ |\\+}
\tl{niśceṣṭo nirvikāraś ca layo jayati yoginā ||\\!}
\end{versinnote}
\end{testimonia}

\begin{philcomm}[hp04_002]
%Adopt alpha reading. [MD: Done.]
The fourth verse quarter of this verse differs from that of the North Indian recension of the \emph{Amanaska} (quoted as the source) but is found in two South Indian manuscripts of the \emph{Amanaska} (Ad and Cc).
\end{philcomm}

%%%%%%%%%%
\subsection*{4.3}
\begin{translation}[hp04_003]
May an extraordinary absorption prevail, in which all volition has been cut off and all activity ceased, intelligible [only] by means of itself and ineffable.%??JM: which is intelligible > intelligible, and other tweaks. Prevail not great but can't think of anything else. 
%% My translation in the Amanaska (reading jāyate): [This] extraordinary absorption by which all volition has been cut off and in which all movement has ceased, is intelligible [only] by means of itself and is beyond the scope of words.
% According to the Amanaska's transmission, jayatāṃ (and its similar readings) arose at a later time.  
\end{translation}

\begin{sources}[hp04_003]
\emph{Amanaska} 2.22
\begin{versinnote}
\tl{ucchinnasarvasaṅkalpo niḥśeṣāśeṣaceṣṭitaḥ |\\+}
\tl{svāvagamyo layaḥ ko 'pi jāyate vāgagocaraḥ ||\\+}
\tl{\var{jāyate ] AllNI (except  Ma Pc), AllSI (except Ad Tha) Ea, jayatāṃ Ma Ad Nb Eb, jayatā Pc, jñāyatāṃ Tha, jagatāṃ AllN (except Nb Nu Ea Eb), japatāṃ Ca, layatāṃ Nu}\\!}
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_003]
\emph{Yogacintāmaṇi} f.~27v (attr.~to the \textit{Rājayoga})
\begin{versinnote}
\tl{ucchinnasarvasaṃkalpo nirgatāśeṣaceṣṭakaḥ |\\+}
\tl{svāvagamyo layaḥ ko 'pi jayatāṃ vāgagocaraḥ ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_003]
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.4}
\begin{translation}[hp04_004]
Dissolution, which is eternal in the elements and senses, takes place where the gaze is. The gaze that has dissolved into the focus becomes the power of living beings.%%??JM: changed meditative focal point to focus (as tr. in v.6)
%JB 'occurs' for 'takes place'
\end{translation}

\begin{sources}[hp04_004]
\emph{Candrāvalokana} 8cd–9ab
\begin{versinnote}
\tl{yatra dṛṣṭir manas tatra bhūtendriyasanātanaṃ |\\+}
\tl{sā śaktis sarvabhūtānāṃ dṛṣṭir lakṣye layaṃ gatā ||\\+}
\tl{\var{sā śaktis ] 4340 4345 T00788, sa śāntis  7970}\\!}
\end{versinnote}

\emph{Kaulajñānanirṇaya} 3.2
\begin{versinnote}
\tl{yatra dṛṣṭir manas tatra bhūtendriya sapudgalaḥ \\+} 
\tl{svaśaktijīvabhūtā hi dṛṣṭilakṣair layaṃ gatāḥ \\+}
\tl{°bhūtā hi ] conj. Hatley, °bhūtāni ABCed\\!} 
% ?? Consult Shaman's critical edition.
\end{versinnote}
\end{sources}

\begin{testimonia}[hp04_004]
\emph{Yogacintāmaṇi} f.~24r
\begin{versinnote}
\tl{rājayoge—\\+}
\tl{yatra dṛṣṭir layas tatra bhūtendriyasanātanī |\\+}
\tl{syāc chaktiḥ sarvabhūtānāṃ dṛṣṭir lakṣeṇa saṃgatā ||\\!}
\end{versinnote}

\emph{Haṭhasaṅketacandrikā} f.~128r
\begin{versinnote}
\tl{tad uktaṃ haṭhapradīpikāyām– \\+}
\tl{yatra dṛṣṭir layas tatra bhūtendriyasanātanī |\\+}
\tl{syāc chaktiḥ sarvabhūtānāṃ dṛṣṭir lakṣeṇa saṃgatā ||\\+}
\tl{\var{syāc chaktiḥ sarva°] B220, syārā śakti serva° 2244}\\+}
\tl{\var{lakṣeṇa saṃgatā ] 2244, lakṣyakṣayaṃ gatā B220}\\!}
\end{versinnote}

\emph{Haṭhatattvakaumudī} 49.25
\begin{versinnote}
\tl{tathā coktaṃ haṭhapradīpikāyām –\\+}
\tl{yatra dṛṣṭir layas tatra bhūtendriyasanātanaḥ |\\+}
\tl{na sā śaktir bhāvabhūtānāṃ dṛṣṭe lakṣe kṣayaṃ gatā ||\\!}
\end{versinnote}
\end{testimonia}

%\begin{philcomm}[hp04_004]
%Adopt dṛṣṭir lakṣye layaṃ gatā. [MD: Done.]
%PYŚ: 2.18 on bhūtendriya and drś
%\end{philcomm}

%%%%%%%%%%
\subsection*{4.5}
\begin{translation}[hp04_005]
The Vedas, Shastras and Puranas are like common prostitutes. \emph{Śāmbhavī mudrā} is unique and guarded like a woman of a good family.
\end{translation}
